\chapter{Pythagorean Theorem}

Watch's Khan Academy's Intro to the Pythagorean Theorem video at \url{https://youtu.be/AA6RfgP-AHU}.

If you have a right triangle, the edges that touch the right angle are
called \emph{the legs}.  The third edge, which is always the longest,
is known as \emph{the hypotenuse}. The Pythagorean Theorem gives us
the relationship between the length of the legs and the length of the
hypotenuse.

\begin{tikzpicture}[scale=1.2]
  \coordinate [circle, fill, inner sep=1pt] (a) at (0,0) ;
  \coordinate [circle, fill, inner sep=1pt] (b) at (0,4) ;
  \coordinate [circle, fill, inner sep=1pt] (c) at (3,0) ;
  \draw (a)--node [outer sep=3pt, left]{Length $a$}(b);
  \draw (b)--node[outer sep=3pt, right]{Length $c$}(c) ;
  \draw (c)--node[outer sep=3pt, below]{Length $b$}(a) ;
  \pic [draw, angle eccentricity=1.5] {right angle = c--a--b};
\end{tikzpicture}

The Pythagorean Theorem tells us that $a^2 + b^2 = c^2$.\index{Pythagorean theorem}

For example, if one leg has length 3 and the other has length 4, then
$a^2 + b^2 = 3^2 + 4^2 = 25$. Thus $c^2$ must equal 25. So you know
the hypotenuse must be of length 5.

(In reality, it almost never works out to be such a tidy number. For
example, what is the length of the hypotenuse if the two legs are 3
and 6? $a^2 + b^2 = 3^2 + 6^2 = 45$.  The length of the hypotenuse is
the square root of that: $\sqrt{45} = \sqrt{9 \times 5} = 3 \sqrt{5}$,
which is approximately 6.708203932499369.)

\begin{Exercise}[title={Find the Missing Length}, label=missingsides]
  What is the missing measure?
  \begin{multicols}{2}
Leg 1 = 6, Leg 2 = 8, Hypotenuse = ? \\(It should be a whole number.)

Leg 1 = 5, Leg 2 = ?, Hypotenuse = 13 \\(It should be a whole number.)
  
Leg 1 = ?, Leg 2 = 15, Hypotenuse = 17 \\(It should be a whole number.)

Leg 1 = 3, Leg 2 = 3, Hypotenuse = ? \\(It is an irrational number. Give the exact answer and then use a calculator to get an approximation.)
\end{multicols}
\end{Exercise}
\begin{Answer}[ref=missingsides]
  10 because $6^2 + 8^2 = 10^2$

  12 because $5^2 + 12^2 = 13^2$

  8 because $8^2 + 15^2 = 17^2$

  $3\sqrt{2} \approx 4.24$ because $3^2 + 3^2 = \left(3 \sqrt{2}\right)^2$
\end{Answer}


\section{Distance between Points}

What is the distance between these two points?\index{distance using Pythagorean theorme}

\begin{tikzpicture}
  % axis
  \draw[thick, <->] (0, -4.1) -- (0, 4.1);
  \draw[thick, <->] (-4.1, 0) -- (4.1, 0);
  \coordinate [circle, fill, inner sep=2pt] (a) at (-1,-2) ;
  \coordinate [circle, fill, inner sep=2pt] (b) at (1,3) ;
  \node [left] at (a) {(-1,-2)};
  \node [right] at (b) {(1,3)};
  % grid
  \draw[help lines, step = 1cm] (-4, -4) grid (4, 4);
  
\end{tikzpicture}

We can draw a right triangle and use the Pythagorean Theorem:

\begin{tikzpicture}
  % axis
  \draw[thick, <->] (0, -4.1) -- (0, 4.1);
  \draw[thick, <->] (-4.1, 0) -- (4.1, 0);
  \coordinate [circle, fill, inner sep=2pt] (a) at (-1,-2) ;
  \coordinate [circle, fill, inner sep=2pt] (b) at (1,3) ;
    \node [left] at (a) {(-1,-2)};
  \node [right] at (b) {(1,3)};
  \coordinate (c) at (1, -2);

  \draw [thick] (a) -- (b);
  \draw [thick] (a) -- node[outer sep = 3pt, below]{Length 2}(c);
  \draw [thick] (c) -- node[rotate=90, outer sep = 3pt, below]{Length 5}(b);
  \draw[help lines, step = 1cm] (-4, -4) grid (4, 4);
 
\end{tikzpicture}


The distance between the two points is $\sqrt{2^2 + 5^2} = \sqrt{29}
\approx 5.385165$. That is, you square the change in $x$ and add it to
the square of the change in $y$. The distance is the square root of
that sum.

\section{Distance in 3 Dimensions}

What if the point is in three dimensional space?  That is, you move 2
meters East, 8 meters North, and 4 meters up in the air. How far are
you from where you started?  You just square each, sum them and take the square root:
$\sqrt{2^2 + 8^2 + 4^2} = \sqrt{84} = 2\sqrt{21} \approx 9.165$ meters.\index{distance!in 3 dimensions}

\begin{tikzpicture}
  \draw [thick, ->] (0,0,0) -- (9,0,0) node[outer sep = 1pt, right]{North} ; 
  \draw [thick, ->]  (0,0,0) -- (0,3,0) node[outer sep = 1pt, above]{Up} ; 
  \draw [thick, ->] (0,0,0) -- (0,0,4) node[outer sep = 1pt, below]{East} ; 

    \draw [dashed]  (8,0,0) -- node[outer sep = 1pt, right]{2}  (8,0,2); 

  \draw [dashed]  (0,0,2) -- node[outer sep = 1pt, below]{8} (8,0,2); 
  \draw [dashed]  (8,0,2) -- node[outer sep = 1pt, right]{4} (8,4,2); 
  \draw [thick]  (0,0,0) --  (8,4,2) node[circle, fill, inner sep=2pt]{}; 
  \node [left] at (5, 2.7, 1){$\sqrt{2^2 + 8^2 + 4^2} \approx 9.165$};

\end{tikzpicture}

