\chapter{Work and Energy}

In this chapter, we are going to talk about how engineers define work
and energy.  We have already talked about force.  We measure force in
newtons, and one newton is the force necessary to accelerate on
kilogram at a rate of one meter per second per second.

When you lean on a wall, you are exerting a force on the wall, but you
aren't doing any work. On the otherhand, if you push a car for a mile,
you are clearly doing work. Work, to an engineer, is the force you
apply to something and the distance that the it moves in the direction
of that force. We measure work in \textit{joules}. A joule is one
newton of force over one meter.

For example, if you push a car uphill with a force of 10 newtons for 12
meters, you have done 120 joules of work.\index{work}

Work is how energy is transferred from one thing to another. When you
push the car, you burn the sugars in your blood. That energy is
transferred to the car: after it has been pushed uphill, the car has
more potential energy.

Thus, we measure the energy something consumes or generates in the
units of work: joules, killowatt hours, horsepower-hours, foot-pounds,
BTUs, and calories.

Let's go over a few different forms that energy can take.

\section{Heat}\index{heat}

When you heat something, you are transferring energy to it.  The BTU
(British Thermal Unit) is a common unit for heat: One BTU is the
amount of heat required to raise the temperature of one pound of water
by one degree. A BTU is about 1,055 joules. When you buy and sell
natural gas as a fuel, it is priced by the BTU.\index{heat} \index{BTU}

\section{Electricity}\index{electricity}

Electricity is the motion of electrons. When you push electrons
through something that resists their passage (like a light bulb),
energy is transferred from the power source (a generator or a battery)
and into the source of the resistance. A \textit{coulomb} is about
$6.24 \times 10^18$ electrons.  If you push one coulomb per second,
you are said to be producing a current of one \textit{ampere} or
\textit{amp}. Electrical resistance is measured in \textit{ohms}.\index{coulombs}\index{amperes or amps}\index{ohms}

\begin{mdframed}[style=important]
The power consumed by a resistance is given by

$$P = I^2R$$

where $P$ is the power consumed in watts, $I$ is the current in amps,
and $R$ is the resistance in ohms.\index{watts}

\end{mdframed}


If your lightbulb consumes 60 watts, and you leave it on for 24 hours,
we say that you have consumed 1.44 kilowatt hours. One kilowatt hour
is the same as 3,600,000 joules.

\section{Chemical Energy}\index{chemical energy}

As mentioned early, some chemical reactions consume energy and some
produce energy. Thus, energy can be stored in the structure of a
molecule. When a plant uses photosynthesis to rearrange water and
carbon dioxide into a sugar molecule, it is converting the energy in
the sunlight into chemical energy. The sugar molecule has more
chemical energy than the carbon dioxide and water molecules that were
used in its creation.

In our diet, we measure this energy in \textit{kilocalories}. A
calorie is the energy necessary to raise one gram of water one degree
Celsius; it is about 4.19 joules. This is a very small unit: an apple
has about 100,000 calories, so people working with food started
measuring everything in kilocalories.\index{calories}

Here is where things get dumb: People who work with food got tired of
saying ``kilocalories'', so they just started using ``Calorie'' to
mean 1,000 calories.  This has created terrible confusion over the
years. If it is capitalized, ``Calorie'' probably means kilocalorie.

\section{Kinetic Energy}\index{kinetic energy}

A mass in motion has energy. For example, if you are in a moving car
and you slam on the breaks, the kinetic energy in the motion of the
car will be converted into heat in the breaks and under the tires.

How much energy does the car have?

\begin{mdframed}[style=important, frametitle={Formula for Kinetic Energy}]

$$E = \frac{1}{2} m v^2$$

where $E$ is the energy in joules, $m$ is the mass in kilograms, and
$v$ is the speed in meters per second.

\end{mdframed}

\section{Gravitational Potential Energy}\index{potential energy!gravitational}

When you lift something heavy onto a shelf, you are giving it
\textit{potential energy}. The amount of energy that you transferred
to it is proportional to its weight and the height that you lifted it.

On the surface of the earth, gravity will accelerate a heavy object at
a rate of $9.8 m/s^2$.

\begin{mdframed}[style=important, frametitle={Formula for Gravitational Potential Energy}]
On earth, then, gravitational potential energy is given by

$$E = (9.8)mh$$


where $E$ is the energy in joules, $m$ is the mass of the object you
lifted, and $h$ is the height that you lifted it.

\end{mdframed}


There are other kinds of potential energy. For example, when you draw
a bow, you have given that bow potential energy. When you release it,
the potential energy is transferred to the arrow, which expresses it
as kinetic energy.

\section{Conservation of Energy}

The first law of thermodynamics says ``Energy is neither created nor
destroyed.''\index{energy!conservation of}

Energy can change forms: Your cells consume chemical energy to give
gravitational potential energy to a car.  However, the total amount of
energy in a closed system stays constant.

\begin{Exercise}[title={The Energy of Falling}, label=energy_falling]
  
A 5 kg cannonball falls off the top of a 3 meter ladder. Just before
it hits the floor, all of its gravitational potential energy has been
converted into kinetic energy.  How fast is the cannonball going when
it hits the floor?

\end{Exercise}
\begin{Answer}[ref=energy_falling]

  At the top of the ladder, the cannonball has $(9.8)(5)(3) = 147$ joules of potential energy.

  At the bottom, the kinetic energy $\frac{1}{2}(5)v^2$ must be equal
  to 147 joules. So $v^2 = \frac{294}{5}$.  Thus it is going about
  $7.7$ meters per second.

  (Yes, a tiny amount of energy is lost to air resistance. For a dense
  object moving at these relatively slow speeds, this energy is
  neglible.)
  
\end{Answer}


\section{Efficiency}

Although energy is always conserved as it moves through different
forms, we aren't always that good at controlling it.\index{efficiency}

For example, an car engine consumes chemical energy in gasoline. Only
about 20\% of the energy consumed is used to turn the wheels.  Most of
the energy is lost as heat. If you run a car for a while, the engine
gets very hot and the exhaust going out the tail pipe is hot.

A human is about 25\% efficient. Most of the loss is in the heat
during the chemical reactions that turn food into motion.

In general, if you are trying to increase efficiency in any system,
the problem is usually easy to identify because heat is produced.

Light bulbs are an interesting case. To get the light of a 60 watt
incandescent bulb, you can use an 8 watt LED or a 16 watt flourescent
light. Thus, we say that the LED light is much more efficient: If you
run both, the incandescent bulb will consume 1.44 kilowatt-hours. The
LED will consume only 0.192 kilowatt hours.

Besides light, the incandescent bulb is producing a lot of heat. If it
is inside your house, what happens to the heat? It warms your house.

In the winter, when you want light and heat, the incandescent bulb is
100\% efficient!

In the summer, if you are running the air conditioner, the
incandescent bulb is worse that just ``inefficent at making light'' --
it is creating heat that has to be removed by the air conditioner! 
