\chapter{Python Lists}

Watch CS Dojo's \textbf{Introduction to Lists in Python} video at \url{https://www.youtube.com/watch?v=tw7ror9x32s}

To review, Python list is an indexed collection. The indices start at
zero. You can create a list using square brackets.

Now you are going to write a program that makes an array of
strings. Type this code into a file called \url{faves.py}:

\begin{Verbatim}
favorites = ["Raindrops", "Whiskers", "Kettles", "Mittens"]
favorites.append("Packages")
print("Here are all my favorites:", favorites)
print("My most favorite thing is", favorites[0])
print("My second most favorite is", favorites[1])
number_of_faves = len(favorites)
print("Number of things I like:", number_of_faves)

for i in range(number_of_faves):
    print(i, ": I like", favorites[i])
\end{Verbatim}

Run it:
\begin{Verbatim}[commandchars=\\\{\}]
$ \textbf{python3 faves.py}
Here are all my favorites: ['Raindrops', 'Whiskers', 'Kettles', 'Mittens', 'Packages']
My most favorite thing is Raindrops
My second most favorite is Whiskers
Number of things I like: 5
0 : I like Raindrops
1 : I like Whiskers
2 : I like Kettles
3 : I like Mittens
4 : I like Packages
\end{Verbatim}
After you have run the code, study it until the output makes sense.

\begin{Exercise}[title={Assign into list}, label=assignintolist]
  Before you list the items, replace "Mittens" with "Gloves".
\end{Exercise}
\begin{Answer}[ref=assignintolist]
\begin{Verbatim}
favorites[3] = "Gloves"
\end{Verbatim}
\end{Answer}

\section{Evaluating Polynomials in Python}

If you had a polynomial like $2x^3 -9x + 12$, you could write it like this: $12x^0 + (-9)x^1 + 0x^2 + 2^x3$.  We could use this representation to keep a polynomial in a Python list:
\begin{Verbatim}
pn1 = [12,-9,0,2]
\end{Verbatim}

In the list, the index of each coefficient would correspond to the degree of that monomial. For example, in the list 2 is at index 3, so that entry represents $2x^3$.

In the last chapter, you evaluated the polynomial $x^3 - 3x^2 + 10x - 12$ at $x=4$. Now you will write code that does that evalution.  Create a file called \url{polynomials.py} and type in the following:

\begin{Verbatim}
def evaluate_polynomial(pn, x):
    sum = 0.0  
    for degree in range(len(pn)):
        coefficient = pn[degree]
        sum = sum + coefficient * x ** degree
        
    return sum
   
pn1 = [-12.0, 10.0, -3.0, 1.0]
y = evaluate_polynomial(pn1, 4.0)
print("Polynomial 1: When x is 4.0, y is", y)
\end{Verbatim}

Run it. It should evaluate to 44.0.

\begin{Exercise}[title={Evaluate Polynomials}, label=pyevalpolynomials]
Using the function that you just wrote, add a few lines of code to \url{polynomials.py} to evaluate the following polynomials:
\begin{itemize}
\item Find $4x^4 - 7x^3 - 2x^2 + 5x + 2.5$ at $x = 8.5$.  It should be 16481.875
\item Find $5x^5 - 9$ at $x = 2.0$.  It should be 151.0
\end{itemize}
\end{Exercise}
\begin{Answer}[ref=pyevalpolynomials]
\begin{Verbatim}
pn2 = [2.5, 5.0, -2.0, -7.0, 4.0]
y = evaluate_polynomial(pn2, 8.5)
print("Polynomia 2: When x is 8.5, y is", y)

pn3 = [-9.0, 0.0, 0.0, 0.0, 0.0, 5.0]
y = evaluate_polynomial(pn3, 2.0)
print("Polynomial 3: When x is 2.0, y is", y)    
\end{Verbatim} 
\end{Answer}


