\chapter{Complex Numbers}

Complex numbers are an extension of the real numbers, which in turn are an extension of the rational numbers. In mathematics, the set of complex numbers is a number system that extends the real number line to a full two dimensions, using the imaginary unit which is denoted by $i$, with the property that $i^2 = -1$.

\section{Definition}

A complex number is a number of the form $a + bi$, where $a$ and $b$ are real numbers, and $i$ is the imaginary unit, with the property that $i^2 = -1$. The real part of the complex number is $a$, and the imaginary part is $b$. 

\section{Why Are Complex Numbers Necessary?}

Complex numbers are essential to many fields of science and engineering. Here are a few reasons why:

\subsection{Roots of Negative Numbers}

In the real number system, the square root of a negative number does not exist because there is no real number that you can square to get a negative number. The introduction of the imaginary unit $i$, which has the property that $i^2 = -1$, allows us to take square roots of negative numbers and gives rise to complex numbers.

\subsection{Polynomial Equations}

The fundamental theorem of algebra states that every non-constant polynomial equation with complex coefficients has a complex root. This theorem guarantees that polynomial equations of degree $n$ always have $n$ roots in the complex plane.

\subsection{Physics and Engineering}

In physics and engineering, complex numbers are used to represent waveforms, in control systems, in quantum mechanics, and many other areas. Their properties make many mathematical manipulations more convenient.

\section{Adding Complex Numbers}

The addition of complex numbers is straightforward. If we have two complex numbers $z_1 = a + bi$ and $z_2 = c + di$, their sum is defined as:

\begin{equation}
z_1 + z_2 = (a + c) + (b + d)i
\end{equation}

In other words, you add the real parts to get the real part of the sum, and add the imaginary parts to get the imaginary part of the sum.

\section{Multiplying Complex Numbers}

The multiplication of complex numbers is a bit more involved. If we have two complex numbers $z_1 = a + bi$ and $z_2 = c + di$, their product is defined as:

\begin{equation}
z_1 \cdot z_2 = (a + bi) \cdot (c + di) = ac + adi + bci - bd = (ac - bd) + (ad + bc)i
\end{equation}

Note the last term comes from $i^2 = -1$. You multiply the real parts and the imaginary parts just as you would in a binomial multiplication, and remember to replace $i^2$ with $-1$.

