\chapter{Naive Bayes Classifier}

The Naive Bayes classifier is a simple yet effective algorithm for
classification tasks. It is a probabilistic classifier based on Bayes'
theorem and some additional simplifying assumptions, which make it
particularly suitable for high-dimensional datasets.\index{naive Bayes
  classifier}

\section{Bayes' Theorem}

Bayes' theorem describes the relationship of conditional probabilities
of statistical quantities. In the context of classification, it can be
written as:

\begin{equation*}
P(C|X) = \frac{P(X|C) \cdot P(C)}{P(X)}
\end{equation*}

where:
\begin{itemize}
\item $P(C|X)$ is the posterior probability of class $C$ given predictor (features) $X$.
\item $P(C)$ is the prior probability of class.
\item $P(X|C)$ is the likelihood which is the probability of predictor given class.
\item $P(X)$ is the prior probability of predictor.
\end{itemize}

\section{The Naivety of Naive Bayes}

The "Naive" in Naive Bayes comes from the assumption that each feature
in the dataset is independent of all other features, given the
class. This is a strong (and often unrealistic) assumption, hence the
name "naive". Despite this unrealistic assumption, the Naive Bayes
classifier often performs well in practice.

In the context of text classification, this naivety translates into
assuming that every word in a document is independent of all other
words, given the document's class.

\section{Working of Naive Bayes Classifier}

When given an instance to classify, the Naive Bayes classifier
calculates the posterior probability of that instance belonging to
each possible class. The classifier then outputs the class with the
highest posterior probability.

For computational reasons, and because the denominator $P(X)$ is
constant given the input, we typically use the following
simplification in practice:

\begin{equation*}
P(C|X) \propto P(X|C) \cdot P(C)
\end{equation*}

which means that we can focus on maximizing $P(X|C) \cdot P(C)$.

Naive Bayes classifiers are highly scalable and are known for their
simplicity, speed, and suitability for high-dimensional datasets.
